Stochastic Variance Reduced Policy Gradient

### 2018-01-20 svrpg.tex
almost finished lemma 1

\begin{restatable}{lemma}{unbias}\label{lemma:unbias}
The SVRPG gradient estimator is unbiased:
\[
\mathop{\mathbb{E}}
\left[\blacktriangledown J(\vtheta)\right] = \gradJ{}.
\]
\end{restatable}

@@

\convergence*
\begin{proof}
We prove the theorem for the following values of the constants:
\begin{align*}
& \psi \coloneqq \min_t\Psi_t, \\
& D \coloneqq \max_t d_t \\
& c_m \coloneqq 0 \\
& c_t \coloneqq c_{t+1}(1+\alpha_t\beta+3\alpha_t^2L^2), \\
\end{align*}
where:
\begin{align*}
	& \beta > 0 \\
	& \Psi_t \coloneqq (\alpha_t - \frac{c_{t+1}\alpha_t}{\beta} - \frac{3}{2}\alpha_t^2L - 3c_{t+1}\alpha_t^2) \\
	& d_t \coloneqq 3(\alpha_t\frac{L}{2} + c_{t+1}\alpha_t^2), \\
\end{align*}
and $\beta,\alpha_t$ are chosen so that $\psi_t > 0$ for $t=1,\dots,m$.
We use two auxiliary lemmas, whose proofs are provided later on:
%Lemma 1
\begin{restatable}[]{lemma}{auxone}\label{lemma:aux1}
\[
	\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \leq
	\frac{R_{t+1}^{s+1} - R_t^{s+1} + Z_t}{\Psi_t},
\]
where
\[
	R_t^{s+1}\coloneqq \Ets{J(\vtheta_t^{s+1}) + c_t\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
\]
\end{restatable}
%Lemma 2
\begin{restatable}[]{lemma}{auxtwo}\label{lemma:aux2}
	...
\end{restatable}

\end{proof}

%Proof of Lemma 1
\auxone*
\begin{proof}
We have:
\begin{align}
	\Ets{J(\vtheta_{t+1}^{s+1})} 
	&\geq \Ets{J(\vtheta_t^{s+1})+\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}} - \frac{L}{2}\norm[]{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}}^2} \\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\gradBlack{\vtheta_t^{s+1}}} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \\
	&\geq \Ets{J(\vtheta_t^{s+1})+\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\gradIdeal{\vtheta_t^{s+1}}+e_s} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}^2} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2}  + \alpha_t\Ets[t-1]{\dotprod{e_s}{\gradJ{\vtheta_t^{s+1}}}} \\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}^2} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} +
	\alpha_t\Covts[t-1]{e_s}{\gradJ{\vtheta_t^{s+1}}} \\
	&\geq  \Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}^2} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} -
	\alpha_t\left|\Covts[t-1]{e_s}{\gradJ{\vtheta_t^{s+1}}}\right| \\
	&\geq  \Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}^2} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \\
	&+
	\alpha_t\frac{\sqrt{\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}}\Varts{\gradJ{\vtheta_t^{s+1}}}}}{\sqrt{N}}
\end{align}
where (1) is from the L-smoothness of $J(\vtheta)$, (2) is from the SVRPG gradient ascent update, (3) is from Definition \ref{def:ideal}, (4) is from the fact that $\EVV[\tau_t\sim\pi(\cdot\mid \vtheta_t^{s+1})]{\gradBlack{\vtheta_t^{s+1}}\mid\vtheta_t^{s+1}} = \EVV[\tau_t\sim\pi(\cdot\mid \vtheta_t^{s+1})]{\gradIdeal{\vtheta_t^{s+1}}\mid\vtheta_t^{s+1}}$, (5) is from the fact that $\Ets[t-1]{e_s} = 0$, and (7)+(8) is from the definition of $\gradApp{\vtheta}{N}$.

>>>>

\begin{restatable}{lemma}{unbias}\label{lemma:unbias}
The SVRPG gradient estimator is unbiased:
\[
\mathop{\mathbb{E}}
\left[\blacktriangledown J(\vtheta)\right] = \gradJ{\vtheta}.
\]
\end{restatable}

@@

%Lemma 2
\begin{restatable}[]{lemma}{auxtwo}\label{lemma:aux2}
	...
\end{restatable}

%Lemma 0
\begin{restatable}[]{lemma}{auxzero}\label{lemma:aux0}
The expected scalar product between $\gradBlack{\vtheta_t^{s+1}}$ and any function $f(\vtheta_t^{s+1})$ which is deterministic for a fixed $\vtheta_t^{s+1}$ can be bounded as follows:
\begin{align*}
\Ets[t]{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} &\geq
\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \\
&\qquad-
\frac{1}{2N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{1}{2}\Ets[t-1]{\norm[]{f(\vtheta_t^{s+1})}^2}
\end{align*}
\end{restatable}
\begin{proof}
\begin{align}
	\Ets{\dotprod{\gradBlack{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}}
	&=
	\Ets{\dotprod{\gradIdeal{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{f(\vtheta_t^{s+1})}} \\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} +
	\Ets[t-1]{\dotprod{e_s}{f(\vtheta_t^{s+1})}} \\
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\dotprod{\Ets{e_s}}{\Ets{f(\vtheta_t^{s+1})}}
	+\Tr\Covts[t-1]{\gradApp{\tilde{\vtheta}^s}{N}}{f(\vtheta_t^{s+1})}  \nonumber\\
	&= 
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad+
	\Tr\Covts[t-1]{\gradApp{\tilde{\vtheta}^s}{N}}{f(\vtheta_t^{s+1})} \\
	&\geq  
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad-
	\Tr\left|\Covts[t-1]{\gradApp{\tilde{\vtheta}^s}{N}}{f(\vtheta_t^{s+1})}\right| \nonumber \\
	%
	&\geq
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad-\Tr\left(\sqrt{\Varts[]{\gradApp{\tilde{\vtheta}^s}{N}}}\circ\sqrt{\Varts[t-1]{f(\vtheta_t^{s+1})}}\right) \nonumber\\
	%
	&\geq	
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad-
	\frac{\alpha_t}{2}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{N}} -\frac{\alpha_t}{2}\Tr\Varts[t-1]{f(\vtheta_t^{s+1})}\\
	%
	&=
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad-
	\frac{1}{2N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{1}{2}\Tr\Varts[t-1]{f(\vtheta_t^{s+1})} \\
	%
	&\geq
	\Ets{\dotprod{\gradJ{\vtheta_t^{s+1}}}{f(\vtheta_t^{s+1})}} \nonumber\\
	&\qquad-
	\frac{1}{2N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{1}{2}\Ets[t-1]{\norm[]{f(\vtheta_t^{s+1})}^2},\nonumber
\end{align}
where (1) is from Definition \ref{def:ideal}; (2) is from the fact that $\gradIdeal{\vtheta_t^{s+1}}$ is both unbiased and independent from $f(\vtheta_t^{s+1})$ \wrt the sampling at time $t$ alone, which is not true for $\gradBlack{\vtheta_t^{s+1}}$; (3) is from $\Ets{e_s}=0$; (4) is from Young's inequality; and (5) is from the definition of $\gradApp{\vtheta}{N}$.
\end{proof}

%Lemma 1
\begin{restatable}[]{lemma}{auxone}\label{lemma:aux1}
The expected squared norm of the true gradient $\gradJ{\vtheta_t^{s+1}}$ can be bounded as follows:
\[
	\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}^2} \leq
	\frac{R_{t+1}^{s+1} - R_t^{s+1} + Z_t}{\Psi_t},
\]
	where
\[
	R_t^{s+1}\coloneqq \Ets{J(\vtheta_t^{s+1}) + c_t\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
\]
\end{restatable}
\begin{proof}
	We have:
	\begin{align}
	\Ets{J(\vtheta_{t+1}^{s+1})} 
	&\geq \Ets{J(\vtheta_t^{s+1})+\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}} - \frac{L}{2}\norm[]{\vtheta_{t+1}^{s+1}-\vtheta_t^{s+1}}^2} \\
	&= \Ets{J(\vtheta_t^{s+1})+\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\gradBlack{\vtheta_t^{s+1}}} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \\
	&\geq
	\Ets{J(\vtheta_t^{s+1})+\alpha_t\norm[]{\gradJ{\vtheta_t^{s+1}}^2} - \frac{\alpha_t^2L}{2}\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2} \nonumber\\
	&\qquad-
	\frac{\alpha_t}{2N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{\alpha_tL^2}{2}\Ets[t-1]{\norm[]{\vtheta_t^{s+1} - \tilde{\vtheta}^s}^2},
\end{align}
where (6) is from the L-smoothness of $J(\vtheta)$, (7) is from the SVRPG update, and (8) is from Lemma \ref{lemma:aux0}.

Next we have:
\begin{align}
\Ets{\norm[]{\vtheta_{t+1}^{s+1}-\tilde{\vtheta}^s}^2} 
&= \Ets{\norm[]{\vtheta_{t+1}^{s+1}- \vtheta_t^{s+1} + \vtheta_t^{s+1}-\tilde{\vtheta}^s}^2} \nonumber\\
&=\Ets{\norm[]{\vtheta_{t+1}^{s+1}-\vtheta_{t}^{s+1}}^2+\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2+2\dotprod{\vtheta_{t+1}^{s+1}-\vtheta_{t}^{s+1}}{\vtheta_t^{s+1}-\tilde{\vtheta}^s}} \nonumber \\
&= \Ets{\alpha_t^2\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2+\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2+2\alpha_t\dotprod{\gradBlack{\vtheta_t^{s+1}}}{\vtheta_t^{s+1}-\tilde{\vtheta}^s}} \\
&= \Ets{\alpha_t^2\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2+\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2+2\alpha_t\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_t^{s+1}-\tilde{\vtheta}^s}} \nonumber\\ 
&\qquad-
\frac{\alpha_t}{N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{1}{2}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2} \\
%
&= \Ets{\alpha_t^2\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2+\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
-2\alpha_t\Ets{\left|\dotprod{\gradJ{\vtheta_t^{s+1}}}{\vtheta_t^{s+1}-\tilde{\vtheta}^s}\right|} \nonumber\\ 
&\qquad-
\frac{\alpha_t}{N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{1}{2}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2} \nonumber\\
%
&\geq \Ets{\alpha_t^2\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2+\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
-2\alpha_t\Ets{\norm[]{\gradJ{\vtheta_t^{s+1}}}\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}} \nonumber\\ 
&\qquad-
\frac{\alpha_t}{N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{1}{2}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2} \nonumber\\
%
&\geq \Ets{\alpha_t^2\norm[]{\gradBlack{\vtheta_t^{s+1}}}^2+\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2}
-2\alpha_t\Ets{\frac{1}{2\beta}\norm[]{\gradJ{\vtheta_t^{s+1}}}^2+\frac{\beta}{2}\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2} \nonumber\\ 
&\qquad-
\frac{\alpha_t}{N}\Tr\Varts[]{\gradApp{\tilde{\vtheta}^s}{1}} -\frac{1}{2}\Ets[t-1]{\norm[]{\vtheta_t^{s+1}-\tilde{\vtheta}^s}^2},
\end{align}
where (9) is from the SVRPG update, (10) is from Lemma \ref{lemma:aux0}, and (11) is from Young's inequality in the 'Peter-Paul' variant.

...[next: put (11),(8) and Lemma A.1 together to finish proof]
\end{proof}

\subsection*{Main theorem}

\convergence*
\begin{proof}
We prove the theorem for the following values of the constants:
\begin{align*}
& \psi \coloneqq \min_t\Psi_t, \\
& D \coloneqq \max_t d_t \\
& c_m \coloneqq 0 \\
& c_t \coloneqq c_{t+1}(1+\alpha_t\beta+3\alpha_t^2L^2), \\
\end{align*}
where:
\begin{align*}
	& \beta > 0 \\
	& \Psi_t \coloneqq (\alpha_t - \frac{c_{t+1}\alpha_t}{\beta} - \frac{3}{2}\alpha_t^2L - 3c_{t+1}\alpha_t^2) \\
	& d_t \coloneqq 3(\alpha_t\frac{L}{2} + c_{t+1}\alpha_t^2), \\
\end{align*}
and $\beta,\alpha_t$ are chosen so that $\psi_t > 0$ for $t=1,\dots,m$.
